Bioverse: Potentially Observable Exoplanet Biosignature Patterns Under the UV Threshold Hypothesis for the Origin of Life

### 2025-02-07 src/tex/ms.tex
trim Bayesian methods

Defining a sample of size $n$ as $X=\{F_\mathrm{NUV, i},L_i\}_{i \in [1,n]}$ where $L_i$ is equal to 1 if life is detected and 0 otherwise, we can calculate the evidence for hypothesis $H_i$ being true against $H_j$ through the Bayes factor
\begin{equation}
BF_{H_i,H_j} = \frac{P(X|H_i)}{P(X|H_j)},
\end{equation}
with $P(X|H_i)$ and $P(X|H_j)$ likelihoods of obtaining the sample $X$ under either hypothesis.

Let $Y=\sum_i^n L_i$ be the random variable counting the number of positive life detections in a sample of size $n$. Its probability mass function under the null hypothesis $H_\mathrm{null}$ is that of a binomial distribution:
\begin{equation}
P(Y=k|H_\mathrm{null}) = \binom{n}{k}f_\mathrm{life}^k(1-f_\mathrm{life})^{n-k}.
\end{equation}

@@

Hence,
\begin{subequations}
\begin{equation}\label{eq:bayes_factor}
BF_{H_1,H_\mathrm{null}} = \frac{P(Y=k|H_1)}{P(Y=k|H_\mathrm{null})} = \frac{\binom{n_\lambda}{k}}{\binom{n}{k}}(1-f_\mathrm{life})^{n_{\lambda}-n},
\end{equation}
and
\begin{equation}\label{eq:bayes_factor2}
BF_{H_\mathrm{null},H_1} = \frac{P(Y=k|H_\mathrm{null})}{P(Y=k|H_1)} = \frac{\binom{n}{k}}{\binom{n_\lambda}{k}}(1-f_\mathrm{life})^{n-n_{\lambda}}.
\end{equation}
\end{subequations}

Given a sample of planets, where for some of them we have convincing biosignature detections but remaining agnostic on $f_\mathrm{life}$: What evidence for $H_\mathrm{1}$ and $H_\mathrm{null}$ can we expect to get?
The analytical expression for the Bayes factor of this inference problem (Equation~\ref{eq:bayes_factor}) is determined by the unknown variables $f_\mathrm{life}$ and $F_\mathrm{NUV, min}$, as well as by the summary statistic $Y$ (number of biosignature detections).
To compute the distribution of evidences, we repeatedly generated samples under $H_\mathrm{1}$ and $H_\mathrm{null}$ and computed the Bayes factors $BF_{H_1,H_\mathrm{null}}$ and $BF_{H_\mathrm{null}, H_1}$.
We then evaluated the fraction of Monte Carlo runs in which certain evidence thresholds~\citep{Jeffreys1939} were exceeded.

@@

>>>>

\rev{Next, we determine the probability distribution of sample outcomes, or likelihood of each hypothesis.}
Let $Y=\sum_i^n L_i$ be the random variable counting the number of positive life detections in a sample of size $n$. Its probability mass function under the null hypothesis $H_\mathrm{null}$ is that of a binomial distribution:
\begin{equation}
    \label{eq:semian:likelihoodHnull}
P(Y=k|H_\mathrm{null}) = \binom{n}{k}f_\mathrm{life}^k(1-f_\mathrm{life})^{n-k}.
\end{equation}

@@

\rev{
Following a similar approach as in \cite{affholder2025interior}, we aim to quantify the information gain from our sampling procedure by computing the Bayes factors~\citep{Jeffreys1939}
}
%\begin{subequations}
\begin{equation}\label{eq:bayes_factor}
BF_{H_1,H_\mathrm{null}} = \frac{P(Y=k|H_1)}{P(Y=k|H_\mathrm{null})} = \frac{\binom{n_\lambda}{k}}{\binom{n}{k}}(1-f_\mathrm{life})^{n_{\lambda}-n}
\end{equation}
and
\begin{equation}
    \label{eq:bayes_factor2}
BF_{H_\mathrm{null},H_1} = \frac{P(Y=k|H_\mathrm{null})}{P(Y=k|H_1)} = \frac{\binom{n}{k}}{\binom{n_\lambda}{k}}(1-f_\mathrm{life})^{n-n_{\lambda}}.
\end{equation}
%\end{subequations}
Given a sample of planets, where for some of them we have convincing biosignature detections but remaining agnostic on $f_\mathrm{life}$: What evidence for $H_\mathrm{1}$ and $H_\mathrm{null}$ can we expect to get?
The analytical expression for the Bayes factor of this inference problem (Equation~\ref{eq:bayes_factor}) is determined by the unknown variables $f_\mathrm{life}$ and $F_\mathrm{NUV, min}$, as well as by the summary statistic $Y$ (number of biosignature detections).
To compute the distribution of evidences, we repeatedly generated samples under $H_\mathrm{1}$ and $H_\mathrm{null}$ and computed the Bayes factors $BF_{H_1,H_\mathrm{null}}$ and $BF_{H_\mathrm{null}, H_1}$.
We then evaluated the fraction of Monte Carlo runs in which certain evidence thresholds~\citep{Jeffreys1939} were exceeded.

@@

\rev{
Our semi-analytical framework focuses on the Bayes ratio as a measure of the evidence that a future observation may bring relative to prior expectations.
This is useful in order to specifically consider the information yield of an observation, but also trades off the opportunity to estimate the posterior probability of a specific hypothesis with the possibility to narrow down the problem to two specific competing hypotheses.
Indeed, the posterior probability depends on the prior expectation for a sample $Y$ (see equation \ref{eq:bayesHone}), which may be approximated as the sum of likelihoods of the modeled hypotheses ($P(Y)=P(Y|H_1)P(H_1)+P(Y|H_{\mathrm{null}})P(H_{\mathrm{null}})$) but could in principle depend on any number of unknown processes.
Could there be factors other than the past UV flux that would determine the occurrence rate of biosignatures in terrestrial exoplanets?
Surely, there is a vast array of possible mechanisms that could be considered which could shape the prior probability of an observed sample of biosignatures $Y$.
Thus, while the Bayes factor is an essential part of the inference process, and the one relevant to prospective analyses of future evidence yields, future attempts at inference may aim for estimating posterior probabilities rather than evidence ratios.
Constraining the prior distribution for the probability that life requires exposure to a certain NUV threshold to emerge is beyond the scope of this study, but remains a question that future work will need to address in order to evaluate fully this hypothesis in the prospect of future observation of biosignatures.}

\rev{Our Bayes factor analysis quantifies how much evidence future data may provide in favor of a hypothesis against another one (the null hypothesis).
Typical frequentist (p-value based) tests usually evaluate only the likelihood of the null hypothesis, thus requiring that the experiment be carefully crafted such that rejection of the null hypothesis can be interpreted in favor of a particular studied hypothesis (for instance that samples from coin tosses score a very low likelihood under the hypothesis of a fair coin may indicate that the coin is biased).
In contrast, Bayes factor analyses such as ours introduce additional constraint to the testing by involving the likelihood of the tested alternate hypothesis (equation \ref{eq:bayes_factor}).
Thus, under Bayes-factor analyses, evidence for the tested hypothesis is not only measured by how unlikely the null hypothesis is given some observation but rather by how much more likely than the null hypothesis is the alternate hypothesis.
The most exhaustive form of hypothesis selection test would be performed by using the entire Bayes theorem (equation \ref{eq:BayesTheorem}), computing the \emph{posterior} probability of the tested hypothesis.
Doing so enables to quantify the information gain from performing some observation (by comparing the posterior with the prior), and to perform inference with consideration for prior information.
\begin{equation}
    \label{eq:BayesTheorem}
    P(H_i|X) = \frac{P(X|H_i)P(H_i)}{P(X)}
\end{equation}
However, implementing the Bayes theorem can prove difficult in practice, owing to the (usually) subjective nature of prior distributions.
Two prior distributions are required for the computation of the posterior distribution: the prior distribution for the considered model and hypothesis, and the prior distribution for the observation.
While the former requires that at least some amount of consensus is identifiable in the community of scientists that discuss this hypothesis, the latter also requires that either empirical distributions for the prior distribution of the observation be available or that the exhaustive list of mechanistic pathways that could potentially lead to this observation be known and tractable.
Taking the ratio of the posterior evidence for two alternate hypotheses (i.e. of equation \ref{eq:BayesTheorem}) allows this problematic term, the prior probability for the observation, to cancel out (equation \ref{eq:postratio}).
\begin{equation}
    \label{eq:postratio}
    \frac{P(H_1|X)}{P(H_0|X)} = \frac{P(X|H_0)P(H_0)}{P(X|H_0)P(H_0)}
\end{equation}
This trades off the opportunity to estimate the posterior probability of a specific hypothesis with the possiblity to narrow down the problem to two specific competing hypotheses.
Could there be factors other than the past UV flux that would determine the occurrence rate of biosignatures in terrestrial exoplanets?
Surely, there is a vast array of possible mechanisms that could be considered which could shape the prior probability of an observed sample of biosignatures $X$.
Truncating the ratio of prior probabilities from equation \ref{eq:postratio} results in the Bayes factor (equation \ref{eq:bayes_factor}), which we used in our semi-analytical analysis (section \ref{sec:results-semianalytical}).
While the Bayes ratio only paints part of the picture of what a full inference pipeline would look line in a future survey aimed at testing origin of life hypotheses from the observation of biosignatures, it captures an essential part of the yield of a particular strategy: by how much would these observations tip the scale of model selection relative to prior expectations.}
